When Imitation Learning Outperforms Reinforcement Learning in Surgical Action Planning

### 2025-07-06 paper_manuscript/results/2025-07-05_21-17-49/fold0/miccai_paper/paper.tex
edit table titles

We evaluate our approaches on the CholecT50 dataset~\cite{nwoye2022cholect50}, which contains 50 laparoscopic cholecystectomy videos with expert-level surgical demonstrations. The dataset provides frame-level annotations for surgical action triplets, consisting of 100 distinct triplet classes combining 6 instruments, 10 verbs, and 15 targets. Each video contains approximately 14,000 frames on average, resulting in a total of approximately 700,000 annotated frames. The dataset represents high-quality expert demonstrations from experienced surgeons performing successful procedures without complications.

@@

\subsection{Evaluation Framework}

\textbf{Temporal Planning Evaluation}: We evaluate planning performance across multiple horizons (1s, 2s, 3s, 5s, 10s, 20s) using mean Average Precision (mAP) computed with the IVT metrics~\cite{nwoye2022data,nwoye2022cholect50}.

@@

\begin{table}[h]
\centering
\caption{Comparative Results: IL vs RL Approaches for Surgical Action Planning. Current refers to t=0s (recognition), Next refers to t+1s prediction.}
\label{tab:main_results}
\begin{tabular}{lcccc}
\toprule
\textbf{Method} & \textbf{Current mAP (\%)} & \textbf{Next mAP (\%)} & \textbf{5s mAP (\%)} & \textbf{10s mAP (\%)} \\
\midrule
DARIL (Ours) & \textbf{34.6} & \textbf{33.6} & \textbf{31.2} & \textbf{29.2} \\
\midrule
DARIL + IRL & \textit{TBD} & \textit{TBD} & \textit{TBD} & \textit{TBD} \\
Latent World Model + RL & \textit{TBD} & \textit{TBD} & \textit{TBD} & \textit{TBD} \\
Direct Video RL & \textit{TBD} & \textit{TBD} & \textit{TBD} & \textit{TBD} \\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{table}[h]
\centering
\caption{Component-wise Performance Analysis of IL Baseline}
\label{tab:component_analysis}
\begin{tabular}{lcc}
\toprule
\textbf{Component} & \textbf{Current mAP (\%)} & \textbf{Next mAP (\%)} \\
\midrule
Instrument (I) & 91.4 & 88.2 \\
Verb (V) & 69.4 & 68.1 \\
Target (T) & 52.7 & 52.5 \\
\midrule
Instrument-Verb (IV) & 42.9 & 38.8 \\
Instrument-Target (IT) & 43.5 & 43.6 \\
Instrument-Verb-Target (IVT) & 34.6 & 33.6 \\
\bottomrule
\end{tabular}
\end{table}

>>>>

We evaluate our approaches on the CholecT50 dataset~\cite{nwoye2022cholect50}, which contains 50 laparoscopic cholecystectomy videos with expert-level surgical demonstrations. The dataset provides frame-level annotations for surgical action triplets, consisting of 100 distinct triplet classes combining 6 instruments, 10 verbs, and 15 targets. Videos are sampled at 1 frame per second, with the training set (40 videos) containing 78,968 frames and the test set (10 videos) containing 21,895 frames, totaling approximately 100,863 annotated frames. The dataset represents high-quality expert demonstrations from experienced surgeons performing full procedures.

@@

\subsection{Evaluation Framework}
1965
\textbf{Temporal Planning Evaluation}: We evaluate planning performance across multiple horizons (1s, 2s, 3s, 5s, 10s, 20s) using mean Average Precision (mAP) computed with the IVT metrics~\cite{nwoye2022data,nwoye2022cholect50}.

@@

\begin{table}[h]
\centering
\caption{Comparative Results: IL vs RL Approaches for Surgical Action Planning. All values are IVT mAP scores in \%. mAP$_{t=0}$ refers to current action recognition, mAP$_{t+1s}$ to 1-second prediction, etc.}
\label{tab:main_results}
\begin{tabular}{lcccc}
\toprule
\textbf{Method} & \textbf{mAP$_{t=0}$} & \textbf{mAP$_{t+1s}$} & \textbf{mAP$_{t+5s}$} & \textbf{mAP$_{t+10s}$} \\
\midrule
DARIL (Ours) & \textbf{34.6} & \textbf{33.6} & \textbf{31.2} & \textbf{29.2} \\
\midrule
DARIL + IRL & \textit{TBD} & \textit{TBD} & \textit{TBD} & \textit{TBD} \\
Latent World Model + RL & \textit{TBD} & \textit{TBD} & \textit{TBD} & \textit{TBD} \\
Direct Video RL & \textit{TBD} & \textit{TBD} & \textit{TBD} & \textit{TBD} \\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{table}[h]
\centering
\caption{Component-wise Performance Analysis of DARIL Baseline. All values are mAP scores in \% for each component and combination.}
\label{tab:component_analysis}
\begin{tabular}{lcc}
\toprule
\textbf{Component} & \textbf{mAP$_{t=0}$} & \textbf{mAP$_{t+1s}$} \\
\midrule
Instrument (I) & 91.4 & 88.2 \\
Verb (V) & 69.4 & 68.1 \\
Target (T) & 52.7 & 52.5 \\
\midrule
Instrument-Verb (IV) & 42.9 & 38.8 \\
Instrument-Target (IT) & 43.5 & 43.6 \\
Instrument-Verb-Target (IVT) & 34.6 & 33.6 \\
\bottomrule
\end{tabular}
\end{table}
